---
title: "Analytical Plan for Non-inferiority of pre-hydrated collagenated xenogenic bone mix in periodontal intrabony defects: clinical trial"
subtitle: 'DOCUMENT: SAP-2023-010-TK-v01'
author: '**From:** Felipe Figueiredo **To:** Tom Kobe'
date: "2023-03-01"
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    reference_docx: misc/style_SAP_en.docx
    toc: yes
# toc-title: "Sumário"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(scipen = 999)
library(pander)
library(knitr)
panderOptions('table.style', 'rmarkdown')
panderOptions('table.split.table', 160)
```

---

\newpage

**Analytical Plan for Non-inferiority of pre-hydrated collagenated xenogenic bone mix in periodontal intrabony defects: clinical trial**

**Document version**

```{r history}
Version <- c("01")
Changes <- c("Initial version")
history <- cbind(Version, Changes)
colnames(history) <- c("Version", "Alterations")
# pander(history, split.cells = 70)
kable(history)
```

```{r cleaning, include=FALSE}
source("scripts/input.R", encoding = 'UTF-8')
```

---

# Abbreviations

<!-- - BMI: body mass index -->
- CAL: clinical attachment level
- CI: confidence interval
<!-- - CV: coefficient of variation -->
- d: Cohen's D effect size
<!-- - HR: hazards ratio -->
<!-- - IQR: interquartile range -->
<!-- - IRR: incidence rate ratio -->
<!-- - OR: odds ratio -->
<!-- - RR: risks ratio -->
- PD: probing depth
- REC: gingival recession
- TP: tip of the papillae
- SD: standard deviation

# Context

## Objectives

To evaluate the non-inferiority of using pre-hydrated collagenated xenogenic bone mix in periodontal intrabony defects.

## Hypotheses

- The CAL change from baseline in Gel 40 is non-inferior to Gen-Os;
- The PD change from baseline in Gel 40 is non-inferior to Gen-Os;
- The REC change from baseline in Gel 40 is non-inferior to Gen-Os;
- The TP change from baseline in Gel 40 is non-inferior to Gen-Os.

# Data

## Raw data

The original data base had
`r Nvar_orig`
variables collected on
`r Nobs_orig`
observations.

<!-- Upon study start the raw data will be collected in a raw table, that will be processed before analysis. -->
<!-- The raw dataset to be collected will have -->
<!-- `r Nvar_orig` -->
<!-- variables collected on -->
<!-- `r Nobs_orig` -->
<!-- observations. -->

<!-- This dataset will include the dates of entry and exit of the cohort, or the date of hospital admission and the date where the endpoint was reached (either an event, or hospital discharge). -->
<!-- Table 1 shows the structure of the raw dataset. -->

<!-- ```{r analytical data} -->
<!-- analytical_mockup %>% -->
<!--   pander(caption = "**Table 1** Raw dataset structure.") -->
<!-- ``` -->

<!-- Each row represents all information collected from each study participant, and each participant included will require a unique study ID. -->
<!-- The outcome should be recorded as a binary variable: either the study participant reached the endpoint -->
<!-- ([outcome]) -->
<!-- or reached the end of study period without experiencing the event. -->
<!-- This information can be recorded in either text form (eg, yes/no), or an indicator -->
<!-- ([outcome] = 1, -->
<!-- end of observation time or discharge = 0). -->
<!-- Exposure and all comorbidities should be recorded as binary variables, whenever possible. -->
<!-- All comorbidities should be recorded as binary variables, whenever possible. -->

<!-- Surgery length should be recorded in -->
<!-- minutes, -->
<!-- hours, -->
<!-- and ICU length in -->
<!-- days. -->
<!-- weeks. -->
<!-- months. -->

<!-- Surgery type will be recorded as a categorical variable. -->
<!-- It is recommended that the number of classes (types of surgery) be as small as possible, to avoid loss of precision in the analysis estimates, without the need to increase the sample size. -->
<!-- Broader categories of surgeries should be preferred, whenever possible. -->

## Analytical dataset

CAL will be calculated as the sum of REC and DP.
All outcomes will be calculated as the difference between the end-of-study measurements and baseline measurements.

After the cleaning process
`r Nvar_final`
variables were included in the analysis
with
`r Nobs_final`
observations.
<!-- After the cleaning process -->
<!-- `r Nvar_final` -->
<!-- variables were included in the analysis. -->
The total number of observations excluded due to incompleteness and exclusion criteria will be reported in the analysis.
Table 1 shows the structure of the analytical dataset.

```{r analytical data}
analytical_mockup %>%
  pander(caption = "**Table 1** Analytical dataset structure after variable selection and cleaning.")
```

All variables in the analytical set were labeled according to the raw data provided and values were labeled according to the data dictionary for the preparation of production-quality results tables and figures.

# Study parameters

## Study design

Randomized controlled clinical trial.

## Inclusion and exclusion criteria

N/A

## Exposures

Efficacy of usage of pre-hydrated collagenated xenogenic bone mix (Gel40) will be tested against the gold standard bone graft (Gen-Os).

## Outcomes

**Specification of outcome measures** (Zarin, 2011):

1. (Domain) Periodontal intrabony defects
2. (Specific measurement) CAL
3. (Specific metric) Change from baseline
4. (Method of aggregation) Mean

**Primary outcome**

CAL change.

**Secondary outcomes**

- REC change;
- PD change;
- TP change.

## Covariates

No adjustment for covariates will be performed in this analysis.

# Statistical methods

## Statistical analyses

### Descriptive analyses

The epidemiological profile of the study participants will be described.
Demographic
<!-- (sex, age and BMI) -->
and clinical variables
<!-- ( [vars] ) -->
will be described as
mean (SD)
<!-- median (IQR) -->
or as counts and proportions (%), as appropriate.
The distributions of participants' characteristics will be summarized in tables and visualized in exploratory plots.

### Inferential analyses

All comparisons between groups will be performed as univariate analyses.
The Cohen's d standardized mean difference and its CI will be calculated for all outcomes as the effect size for the efficacy of the main exposure relative to the gold standard.

The non-inferiority of the test intervention will be determined by using the minimum standardized difference detectable by this study d (calculated in section 5.3), where the equivalence interval lies between -d and d.
For each outcome non-inferiority will be proven if the CI lies entirely above the lower margin of the equivalence interval -d.

The p-values from the
independent t test with Welch correction,
will be reported for information purposes.

<!-- All inferential analyses will be performed in the statistical models (described in the next section). -->

### Statistical modeling

N/A

### Missing data

No missing data imputation will be performed.
All evaluations will be performed as complete case analyses.
Missing data counts and proportions will be reported in tables.

## Significance and Confidence Intervals

All analyses will be performed using the significance level of 5%.
All significance hypothesis tests and confidence intervals computed will be
two-tailed.
<!-- left-tailed. -->
<!-- right-tailed. -->

## Study size and Power

```{r}
n <- analytical %>%
  count(exposure) %>%
  pull(n)
n1 <- n[1]
n2 <- n[2]
total <- sum(n)
d <- pwr::pwr.t.test(n = total/2, power = .8, sig.level = .05)$d %>%
  format.float()
```

Study groups were balanced and the total sample size was `r total` including all groups.

<!-- Study groups were not balanced with -->
<!-- group1 with `r n1` participants -->
<!-- and -->
<!-- group2 with `r n2` participants -->
<!-- with a total study size of `r total`. -->
<!-- This power analysis adjusts for different group sizes. -->

With these group sizes a t test can detect an effect size as large as
**d = `r d`**
with 80% power and 5% significance level (Cohen, 1988).

## Statistical packages

This analysis will be performed using statistical software `R` version `r getRversion()`.

# Observations and limitations

**Small sample size**

The total sample size of this study is only 20 participants.
Cohen considered an effect size of d = 0.8 to be large, which means that this sample size is only able to detect very large clinical effects.

**Recommended reporting guideline**

The adoption of the EQUATOR network (<http://www.equator-network.org/>) reporting guidelines have seen increasing adoption by scientific journals.
All clinical trials are recommended to be reported following the CONSORT guideline (Schulz K F, Altman D G, Moher D., 2010).

# References

- **SAR-2023-010-TK-v01** -- Non-inferiority of pre-hydrated collagenated xenogenic bone mix in periodontal intrabony defects: clinical trial
- Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd Ed.). New York: Routledge.
- Zarin DA, et al. The ClinicalTrials.gov results database -- update and key issues. N Engl J Med 2011;364:852-60 (<https://doi.org/10.1056/NEJMsa1012065>).
- Gamble C, et al. Guidelines for the Content of Statistical Analysis Plans in Clinical Trials. JAMA. 2017;318(23):2337–2343 (<https://doi.org/10.1001/jama.2017.18556>).
- Schulz K F, Altman D G, Moher D. CONSORT 2010 Statement: updated guidelines for reporting parallel group randomised trials BMJ 2010; 340 :c332 (<https://doi.org/10.1136/bmj.c332>).

# Appendix

This document was elaborated following recommendations on the structure for Statistical Analysis Plans (Gamble, 2017) for better transparency and clarity.

## Availability

All documents from this consultation were included in the consultant's Portfolio.

<!-- The client has requested that this analysis be kept confidential until a future date, determined by the client. -->
<!-- All documents from this consultation are therefore not published online and only the title and year of the analysis will be included in the consultant's Portfolio. -->
<!-- After the agreed date is reached, the documents will be released. -->

<!-- The client has requested that this analysis be kept confidential. -->
<!-- All documents from this consultation are therefore not published online and only the title and year of the analysis will be included in the consultant's Portfolio. -->

The portfolio is available at:

<https://philsf-biostat.github.io/SAR-2023-010-TK/>
